Visualizing Independence Using Extended Association Plots

### 2005-08-12 trunk/vcd/paper/JCGS.Rnw
KH and DM changes included, abstract added, only intro missing
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@441 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%% Hyperref
\hypersetup{%
  pdftitle = {Residual-based Shadings for Visualizing (Conditional) Independence},
  pdfsubject = {Residual-based Shadings},
  pdfkeywords = {contingency tables, association plots, mosaic plots, HCL colors, 
    HSV colors, conditional inference, permutation tests},
  pdfauthor = {Achim Zeileis, David Meyer, Kurt Hornik},
  %% change colorlinks to false for pretty printing
  colorlinks = {true},
  linkcolor = {Blue},
  citecolor = {Blue},
  urlcolor = {Red},
  hyperindex = {true},
  linktocpage = {true},
}

@@

hue.slice <- function(hue, grid.n = 100, type = c("HCL", "HSV"), plot = TRUE, fixup = FALSE)
{
  type <- match.arg(type)
  if(type == "HCL") {
    chroma = seq(0, 100, length = grid.n)
    luminance = seq(0, 100, length = grid.n)
    nc <- length(chroma)
    nl <- length(luminance)
    color.slice <- outer(chroma, luminance, function(y, x) hcl(hue, x, y, fixup = fixup))
    xlab <- "chroma"
    ylab <- "luminance"
    main <- paste("hue =", round(hue, digits = 0))
  } else {
    chroma = seq(0, 1, length = grid.n)
    luminance = seq(0, 1, length = grid.n)
    nc <- length(chroma)
    nl <- length(luminance)
    color.slice <- outer(chroma, luminance, function(y, x) hsv(hue, x, y))
    xlab <- "saturation"
    ylab <- "value"
    main <- paste("hue =", round(hue, digits = 3))
  }
  if(plot) {
    plot(0.5, 0.5, xlim = range(chroma), ylim = range(luminance), type = "n", axes = FALSE,
         xlab = xlab, ylab = ylab, yaxs = "i", xaxs = "i", main = main)
    for(i in 1:(nc-1)) {
      rect(chroma[i], luminance[-nl], chroma[i] + 100/(nc-1), luminance[-1], border = color.slice[,i+1], col = color.slice[,i+1])
    }
    axis(1)
    axis(2)
    box()
  }
  colnames(color.slice) <- chroma
  rownames(color.slice) <- luminance
  attr(color.slice, "type") <- type
  class(color.slice) <- "slice"
  invisible(color.slice)
}
@

\begin{abstract}
Visualizing independence by association and mosaic plots.

Enhancements: diverging color palette based on HCL color space
combined with visualization of the result of a significance test.

Extensions to visualizing conditional independence.\\[0.3cm]

@@

@@

Association plots as residual plots for log-linear models
\citep{vcd:Meyer+Zeileis+Hornik:2003}, especially
in coplot or trellis layout.

@@

But the sum of squares is not the only plausible way of capturing deviations
from $0$ in the residuals.
There are many other conceivable functionals $\lambda(\cdot)$ which
lead to reasonable test statistics $\lambda([r_{ij}])$: one which is particularly
suitable for identifying the cells which cause the dependence (if any) is
the maximum of the absolute values
\begin{equation} \label{eq:MaxAbs}
M \quad = \quad \max_{i, j} |r_{ij}|.
\end{equation}
Given a critical value $c_\alpha$ for this test statistic, all residuals whose
absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at significance level
$\alpha$ \cite[][ch.~7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells causing
the dependence can easily be identified.

@@

Other measures of discrepancy (e.g., deviance residuals) could, of course, also be used
instead of the the Pearson residuals $[r_{ij}]$. But as the ideas discussed here
extend straightforwardly to that situation, we do not go into detail about this.

@@

The extensions of \cite{vcd:Friendly:1994} to mosaic plots provide a
substantial improvement of the original mosaic plots enhancing them
from a plot for contingency tables to a visualization technique for
log-linear models---and thus also for independence problems including
2-way tables as the simplest case.

@@

The way the (light) blue and red colors are chosen differs somewhat between
various implementations of Friendly mosaic plots: In his original 
\proglang{SAS} implementation \citep[see][]{vcd:Friendly:2000}, Michael Friendly
uses colors from a palette based on HLS (Hue--Luminance--Saturation) color space.
The implementation in the \proglang{R} system for statistical computing and graphics
\citep{vcd:R:2005} employs colors from HSV (Hue--Saturation--Value) space. 
Both spaces are rather similar transformations of RGB (Red--Green--Blue) space
\citep{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \citep{vcd:Moretti+Lyons:2002} making the generation
of the Friendly shading very simple. As both spaces can generate the
same colors, we only discuss HSV space in the following.

The HSV space originally looks like a cone with black at its
peak (zero value) and full color wheels for different saturations at the other end,
centered around a white (full value). Type and amount of color are controlled by hue
and saturation, respectively. Typically, polar coordinates $(h, s, v)$ rescaled to the 
unit interval are used in this space, giving it the appearance of a cylinder.
For generating colors in the Friendly shading, the following strategy is used:
The hue $h$ codes the sign of the residuals---$h = 0$ (red hue) is used for negative residuals,
$h = 2/3$ (blue hue) for positive residuals. The absolute size of the residuals
is then coded by the saturation $s$ which is set to $0$, $0.5$ and $1$, respectively,
for small/medium/large residuals. The value is always fixed at $v = 1$.
This is also depicted in the upper panel of Figure~\ref{fig:shadingHSVHCL} which shows
two slides from the HSV cylinder side by side. The plot shows the saturation/value plane
for the given hues $h = 0$ and $h = 2/3$. The significant color palette shows the
colors used for the Friendly shading when the residuals are increasing from left to right.
The set of non-significant colors will be explained in Section~\ref{sec:significance}.

@@

Alternative ways to choose colors have been available for a long time, but have
been only slowly adopted for implementations of colors in computer packages in 
general and for shading in statistical graphs in particular. The idea of using
perceptually based colors that are `in harmony' go back to \cite{vcd:Munsell:1905} 
who introduced a color notation for balanced colors. Based on similar principles, Cynthia
Brewer and co-workers suggested different types of palettes (qualitative/sequential/diverging)
and provided the online tool \pkg{ColorBrewer.org} \citep{vcd:Harrower+Brewer:2003} for
selecting an appropriate palette for a specific problem. Furthermore, the
\citet[CIE,][]{vcd:CIE:2004} introduced the two perceptually
based color spaces CIELAB and CIELUV where the latter is typically preferred for 
emissive color technologies such as computer displays. \cite{vcd:Ihaka:2003} discusses
how CIELUV colors can be used for choosing qualitative palettes for statistical graphics such as barplots.
By taking polar coordinates in CIELUV space, it is called HCL (Hue--Chroma--Luminance) space
and qualitative palettes can easily be chosen by using a range of hues for fixed values
of chroma and luminance. Such colors are always balanced towards the same grey 
and thus do not have the problem of varying saturations as the HSV colors.

Here, we discuss how similar ideas can be used for deriving diverging HCL palettes that 
provide a suitable translation of the ideas from the Friendly shading to perceptually
uniform HCL colors. 
The HCL space looks like a distorted double cone with black (zero luminance) at one end and white
at the other (full luminance). In its middle, there is a full color wheel for different
values of chroma (that controls the colorfulness). Unfortunately, the HCL space is not as
regular as the HSV space and consequently cannot be standardized in the same way as the
HSV space. Its dimensions are usually given by a
hue ranging in $[0, 360]$ degrees and chroma and luminance ranging in $[0, 100]$
percent. But not all combinations $(h, c, l)$ yield valid HCL colors and the admissable combinations
of $c$ and $l$ vary across different hues $h$. But for the problem of constructing
a diverging palette, this problem can easily be overcome as we just need two different
hues (a `negative' and a `positive' hue) and hence we can choose two hues that correspond to 
similar shapes in the chroma/luminance plane. The lower panel of Figure~\ref{fig:shadingHSVHCL}
shows two such planes side by side for the hues $h = 0$ and $h = 260$. To obtain a 
sequence of colors with the same properties as the Friendly shading, the palette starts
at a fully saturated red $(0, 100, 50)$, goes via a neutral color, ends at a fully saturated
blue $(260, 100, 50)$, and uses linear interpolation in between. Instead of using white
$(0, 0, 100)$ as the neutral color, a light grey $(0, 0, 90)$ is employed as motivated above.

@@

lines(c(-18.75, -11.25), c(52.5, 67.5), lty = 2)
lines(c(-8.75, -1.25), c(72.5, 87.5), lty = 2)
lines(c(18.75, 11.75), c(52.5, 67.5), lty = 2)
lines(c(8.75, 1.25), c(72.5, 87.5), lty = 2)
@
\caption{Extended shading in HSV (upper) and HCL space (lower).}
\label{fig:shadingHSVHCL}
\end{center}
\end{figure}

@@

The shading scheme of \cite{vcd:Friendly:1994} was suggested to visualize
the pattern of dependence in contingency tables, as discussed above, but
the presence (or absence) of colors in a plot also always conveys an
impression of interestingness (or un-interstingness, respectively). That is,
viewers might be tempted to interpret the absence of color in a plot
as a clue that there is no significant departure from independence. Or vice
versa, colored cells would convey the impression that there is significant
dependence. Currently, both is not true as can be seen in the left panel
of Figure~\ref{fig:shadings} which shows the mosaic display for the arthritis
data with Friendly shading. Although there is significant dependence, no 
residual exceeds an absolute value of $2$ and hence no cell is colored.
Of course, it can be argued that the shading was not designed for this purpose
and that different cut offs than $2$ and $4$ should be used here. But then
again, it would be nice if these cut offs could be chosen automatically in a
data-driven fashion which is what we do below.

The Friendly shading can be interpreted to be a visualization of the maximum
statistic $M$ from Equation~(\ref{eq:MaxAbs}) which always employs the critical values $c_\alpha$
$2$ and $4$. But the problem is that it is not clear to which
significance levels $\alpha$ these critical values correspond because the distribution
of $M$ depends on the underlying contingency table. The natural solution to this
problem is to compute the critical values from the distribution of $M$ in a data-driven
way (i.e., for the table visualized) and use these instead of the hard-coded values
$2$ and $4$. In the right panel of Figure~\ref{fig:shadings} this is done for
the arthritis data by employing the critical values
$\Sexpr{round(as.vector(art_max$qdist(0.9)), digits = 2)}$ at level $\alpha = 0.1$ and
$\Sexpr{round(as.vector(art_max$qdist(0.99)), digits = 2)}$ at level $\alpha = 0.01$
(and using the diverging HCL palette) derived from the permutation distribution of $M$ for 
the arthritis data as described in Section~\ref{sec:tests}. By using these cut offs, the presence of color
in the plot is equivalent to significance (of the maximum statistic $M$) at 
level $\alpha = 0.1$ and $\alpha = 0.01$, respectively, and exactly the cells which `cause'
the dependence are highlighted. For the arthritis data, these are in particular the 
cells in the last column that signal that there are significantly more
treated patients and fewer placebo patients with marked improvement than would be expected
under independence.

@@

This maximum shading is already very flexible and combines visualization and inference.
However, it can only be applied when employing the maximum statistic because it is the only
aggregation functional $\lambda(\cdot)$ where a single large residual $|r_{ij}|$ exceeding
its critical value is equivalent
to a significant value of the whole test statistic $\lambda([r_{ij}])$.\footnote{This
can easily be seen from the following consideration: To obtain a level $\alpha$ test, where 
any residual $|r_{ij}|$ exceeding the critical value $c_\alpha$ gives a significant test
result, the equation $\alpha = P(\mbox{any } |r_{ij}| > c_\alpha)$ must hold. This is equivalent
to $1 - \alpha = P(\mbox{all } |r_{ij}| \le c_\alpha) = P(\max_{i,j} |r_{ij}| \le c_\alpha)$.}
Typically, applying the maximum statistic is feasible and also appropriate for exploratory 
analysis, but it would be desirable to also have a residual-based shading that can incorporate
visualization of significance when the sum-of-squares statistic $X^2$ (or any other functional
$\lambda(\cdot)$) is used. For the reasons discussed above, it is not possible to achieve this by shading
individual cells differently but can only be realized by using different colors for the
whole table. As outlined before, colorfulness is intuitively matched with interestingness,
therefore a rather natural idea is to use the fully colored palette only when the corresponding
test is significant and to use a less colorful palette if not. For the HCL scheme, the
amount of color can conveniently be controlled by varying the maximum chroma value used. For the full
colors, the maximum chroma was set to $100$ as shown in Figure~\ref{fig:shadingHSVHCL} and 
is reduced to $20$ for the non-significant palette. This palette still codes the absolute
size of the residuals by luminance (i.e., the amount of grey), uses the same neutral grey
for small `un-interesting' residuals, codes positive and negative residuals by different hues,
but gives less emphasis to the pattern by making the plot less colorful. A similar effect
can be obtained in the HSV space if the value is reduced from $1$ to $0.5$ for non-significant 
palettes (see Figure~\ref{fig:shadingHSVHCL}). But as for the significant colors,
the dimensions used for creating this palette are confounded and hence the HCL scheme is clearly
preferable.

@@

To introduce the new residual-based shadings without too much overhead in
Section~\ref{sec:shadings}, we have only considered the independence problem
in 2-way tables. But, in fact, none of the ideas described are really
confined to the 2-dimensional case as will be outlined in this section.

Mosaic displays have been emphasized in the literature to be an excellent means
of visualization for log-linear models \citep{vcd:Friendly:1999,vcd:Theus+Lauer:1999}.
Problems of interest include complete, joint or conditional independence for all of 
which the tables of expected values and residuals (again Pearson or deviance) can
be computed. \cite{vcd:Friendly:1994,vcd:Friendly:1999} shows that his residual-based 
shading scheme can directly be applied to these more complex independence models.
For inference, the most commonly used aggregation functional for the residuals
is again the sum of squares yielding the associated Pearson or likelihood
ratio statistic, respectively \citep{vcd:Agresti:2002}. Hence, our residual-based
shading with different HCL palettes for significant and non-significant results
can directly be applied to these models. For the maximum shading with
data-driven cut offs, even more flexibility is possible than in the 2-way case
when the structure of the independence problem is exploited which will be discussed 
below for conditional independence

@@

To show how the structure of the independence problem can be exploited for 
choosing appropriate critical values in the maximum shading and for selecting 
a suitable layout of mosaic or association plots, we consider the conditional
independence problem. To keep the notational overhead low again, only the 3-way
case is spelt out, but the same ideas extend fairly straightforwardly to conditional
independence in higher dimensions. In a 3-way table $[n_{ijk}]$ with underlying
theoretical probability distribution $[p_{ijk}]$ the conditional independence
problem can be formulated as
\begin{equation} \label{eq:H0ci}
H_0: \; \pi_{ij|k} \quad = \quad \pi_{i+|k} \pi_{+j|k}.
\end{equation}
where $\pi_{ij|k}$ are the conditional probabilities given the stratum $k$
with $k = 1, \dots, K$. Under the assumption of conditional independence,
we can again compute expected frequencies $[\hat n_{ijk}]$ and the corresponding 
residuals $[r_{ijk}]$. To test the conditional independence hypothesis, usually
the sum-of-squares statistic is used
\begin{equation} \label{eq:sumChisq}
\sum_{i, j, k} r_{ijk}^2 \quad = \quad \sum_k X_k^2,
\end{equation}
which is simply the sum of the individual sum-of-squares statistics $X_k^2$ in each stratum $k$.

@@

All these statistics are of type
\begin{equation} \label{eq:coindep_test}
\lambda_\mathrm{agg} (\lambda_\mathrm{indep}(r_{ijk})),
\end{equation}
where $\lambda_\mathrm{indep}$ is a functional for assessing independence in stratum $k$
and $\lambda_\mathrm{agg}$ is a functional for aggregating over the $k = 1, \dots, K$ strata.
If the maximum is used for the latter, then identification of the strata causing the
independence is possible. If additionally $\lambda_\mathrm{indep}$ is the maximum,
the corresponding cells can also be identified. Hence, the double maximum
statistic from Equation~(\ref{eq:doublemax}) is the only functional allowing for detection of
both the strata and the cells which caused the dependence.

But the main purpose of the formulation of the different test statistics is not so much 
inference but their applicability to diagnostic plots via residual-based shadings. As already
discussed, it is possible for all aggregation functionals to simply use either
the significant or the non-significant shading for all cells in the contingency table---this
strategy would have to be used for the sum-of-squares statistic from Equation~(\ref{eq:sumChisq}). If
$\lambda_\mathrm{agg}$ is the maximum as in Equation~(\ref{eq:maxChisq}),
then the significant palette would only be used
in those strata in conflict with the hypothesis of conditional independence whereas the
non-significant palette would be used for the remaining strata. Finally, if both $\lambda_\mathrm{agg}$
and $\lambda_\mathrm{indep}$ are the maximum, then the same strategy as in Section~\ref{sec:shadings}
can be pursued, i.e., only the significant palette is used but with data-driven cut offs
derived from the distribution of the double maximum statistic from Equation~(\ref{eq:doublemax}).

@@

Various strategies for constructing residual-based shadings for visualizing
(conditional) independence in contingency tables via mosaic and association plots
are discussed. The shading
of \cite{vcd:Friendly:1994} is extended in two directions: the use of perceptually
uniform HCL colors and the combination of visualization and significance testing.
To achieve the former, a general guideline for constructing diverging palettes
in HCL space is introduced. The advantages of using this HCL shading scheme
instead of an HSV scheme are that the colors from this perceptually based color space
are device independent and provide uniform saturations over different hues.
Furthermore, the colorfulness in this shading can be controlled independently from
the other two dimensions.

>>>>

%% Hyperref
\hypersetup{%
  pdftitle = {Residual-based Shadings for Visualizing (Conditional) Independence},
  pdfsubject = {Residual-based Shadings},
  pdfkeywords = {association plots, conditional inference, contingency tables, HCL colors, HSV colors, mosaic plots},
  pdfauthor = {Achim Zeileis, David Meyer, Kurt Hornik},
  %% change colorlinks to false for pretty printing
  colorlinks = {true},
  linkcolor = {Blue},
  citecolor = {Blue},
  urlcolor = {Red},
  hyperindex = {true},
  linktocpage = {true},
}

@@

hue.slice <- function(hue, grid.n = 101, type = c("HCL", "HSV"), plot = TRUE, fixup = FALSE)
{
  type <- match.arg(type)
  if(type == "HCL") {
    chroma = seq(0, 100, length = grid.n)
    luminance = seq(0, 100, length = grid.n)
    nc <- length(chroma)
    nl <- length(luminance)
    color.slice <- outer(chroma, luminance, function(y, x) hcl(hue, x, y, fixup = fixup))
    xlab <- "chroma"
    ylab <- "luminance"
    main <- paste("hue =", round(hue, digits = 0))
  } else {
    chroma = seq(0, 1, length = grid.n)
    luminance = seq(0, 1, length = grid.n)
    nc <- length(chroma)
    nl <- length(luminance)
    color.slice <- outer(chroma, luminance, function(y, x) hsv(hue, x, y))
    xlab <- "saturation"
    ylab <- "value"
    main <- paste("hue =", round(hue, digits = 3))
  }
  if(plot) {
    plot(0.5, 0.5, xlim = range(chroma), ylim = range(luminance), type = "n", axes = FALSE,
         xlab = xlab, ylab = ylab, yaxs = "i", xaxs = "i", main = main)
    for(i in 1:(nc-1)) {
      rect(chroma[i], luminance[-nl], chroma[i] + 100/(nc-1), luminance[-1], border = color.slice[,i+1], col = color.slice[,i+1])
    }
    axis(1)
    axis(2)
    box()
  }
  colnames(color.slice) <- chroma
  rownames(color.slice) <- luminance
  attr(color.slice, "type") <- type
  class(color.slice) <- "slice"
  invisible(color.slice)
}
@

\begin{abstract}
Residual-based shadings for enhancing mosaic and association plots
to visualize independence models for contingency tables are extended
in two directions: (a)~perceptually uniform HCL colors are used and (b)~the
result of an associated significance test is coded by the presence of
color in the visualization. For obtaining (a), a general strategy for
deriving diverging palettes in the perceptually-based HCL space is suggested. As for (b),
cut offs that control the appearance of color are computed in a data-driven
way based on the conditional permutation distribution of maximum-type
test statistics. The shadings are first established for the case of independence
in 2-way tables and then extended to more general independence models
for multi-way tables, including in particular the conditional independence
problem.\\[0.3cm]

@@

explore independence structure in contingency tables by log-linear models and visually
by mosaic displays

@@

@@

However, the sum of squares is not the only plausible way of capturing deviations
from $0$ in the residuals.
There are many other conceivable functionals $\lambda(\cdot)$ which
lead to reasonable test statistics $\lambda([r_{ij}])$: one which is particularly
suitable for identifying the cells responsible for the dependence (if any) is
the maximum of the absolute values
\begin{equation} \label{eq:MaxAbs}
M \quad = \quad \max_{i, j} |r_{ij}|.
\end{equation}
Given a critical value $c_\alpha$ for this test statistic, all residuals whose
absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at significance level
$\alpha$ \cite[][ch.~7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells responsible
for the dependence can easily be identified.

@@

Note, that virtually all ideas discussed in this paper also extend straightforwardly to the situation
where other measures of discrepancy (such as, e.g., deviance residuals) are used
instead of the Pearson residuals $[r_{ij}]$.

@@

The extensions of \cite{vcd:Friendly:1994} to mosaic plots provide a
substantial improvement of the original mosaic plots enhancing them
from a plot for contingency tables to a visualization technique for
log-linear models and their residuals---and thus also for independence problems including
2-way tables as the simplest case.

@@

The way the (light) blue and red colors are chosen differs somewhat between
various implementations of Friendly mosaic plots: In his original 
\proglang{SAS} implementation \citep[see][]{vcd:Friendly:2000}, Michael Friendly
uses colors from a palette based on HLS (Hue-Luminance-Saturation) color space.
The implementation in the \proglang{R} system for statistical computing and graphics
\citep{vcd:R:2005} employs colors from HSV (Hue-Saturation-Value) space. 
Both spaces are rather similar transformations of RGB (Red-Green-Blue) space
\citep{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \citep{vcd:Moretti+Lyons:2002} making the generation
of the Friendly shading very simple. As both spaces can generate the
same colors, we only discuss HSV space in the following.

The HSV space originally looks like a cone\footnote{The wikipedia gives a very useful
overview at \url{http://en.wikipedia.org/wiki/HSV_color_space} which also provides
links to comparisons with other color spaces discussed in this paper.}
with black at its peak (zero value) and full color wheels for different saturations at the other end,
centered around a white (full value). Type and amount of color are controlled by hue
and saturation, respectively. Typically, polar coordinates $(h, s, v)$ rescaled to the 
unit interval are used in this space, giving it the appearance of a cylinder.
For generating colors in the Friendly shading, the following strategy is used:
The hue $h$ codes the sign of the residuals---$h = 0$ (red hue) is used for negative residuals,
$h = 2/3$ (blue hue) for positive residuals. The absolute size of the residuals
is then coded by the saturation $s$ which is set to $0$, $0.5$ and $1$, respectively,
for small/medium/large residuals. The value is always fixed at $v = 1$.
This is also depicted in the upper panel of Figure~\ref{fig:shadingHSVHCL} which shows
the saturation/value plane for the given hues $h = 0$ and $h = 2/3$. 
The significant color palette shows the
colors used for the Friendly shading when the residuals are increasing from left to right.
The set of non-significant colors will be explained in Section~\ref{sec:significance}.

@@

Alternative ways to choose colors have been available for a long time, but have
been only slowly adopted for implementations of colors in computer packages in 
general and for shading in statistical graphs in particular. The idea of using
perceptually-based colors that are `in harmony' go back to \cite{vcd:Munsell:1905} 
who introduced a color notation for balanced colors. Based on similar principles, Cynthia
Brewer and co-workers suggested different types of palettes (qualitative/sequential/diverging)
and provided the online tool \pkg{ColorBrewer.org} \citep{vcd:Harrower+Brewer:2003} for
selecting an appropriate palette for a specific problem. Furthermore, the
\citet[CIE,][]{vcd:CIE:2004} introduced the two perceptually
based color spaces CIELAB and CIELUV where the latter is typically preferred for 
emissive color technologies such as computer displays. \cite{vcd:Ihaka:2003} discusses
how CIELUV colors can be used for choosing qualitative palettes for statistical graphics such as barplots.
By taking polar coordinates in CIELUV space, it is called HCL (Hue-Chroma-Luminance) space
and qualitative palettes can easily be chosen by using a range of hues for fixed values
of chroma and luminance. Such colors are always balanced towards the same grey 
and thus do not have the problem of varying saturations as the HSV colors.

Here, we discuss how similar ideas can be used for deriving diverging HCL palettes that 
provide a suitable translation of the ideas from the Friendly shading to perceptually
uniform HCL colors. 
The HCL space looks like a distorted double cone with black (zero luminance) at one end and white
at the other (full luminance). In its middle, there is a full color wheel for different
values of chroma (that controls the colorfulness). Unfortunately, the HCL space is not as
regular as the HSV space and consequently cannot be standardized in the same way as the
HSV space. Its dimensions are usually given by a
hue ranging in $[0, 360]$ degrees and chroma and luminance ranging in $[0, 100]$
percent. However, not all combinations $(h, c, l)$ yield valid HCL colors and the admissable combinations
of $c$ and $l$ vary across different hues $h$. For the task of constructing
a diverging palette, this problem can easily be overcome as we just need two different
hues (a `negative' and a `positive' hue) and hence we can choose two hues that correspond to 
similar shapes in the chroma/luminance plane. The lower panel of Figure~\ref{fig:shadingHSVHCL}
shows two such planes side by side for the hues $h = 0$ and $h = 260$.\footnote{The hue
$h = 260$ is chosen (rather than $h = 240$ which would correspond to $h = 2/3$ in HSV space)
because its chroma/luminance plane is most similar to that of $h = 0$.}
To obtain a 
sequence of colors with the same properties as the Friendly shading, the palette starts
at a fully saturated red $(0, 100, 50)$, goes via a neutral color, ends at a fully saturated
blue $(260, 100, 50)$, and uses linear interpolation in between. Instead of using white
$(0, 0, 100)$ as the neutral color, a light grey $(0, 0, 90)$ is employed as motivated above.

@@

lines(c(-18.75, -11.25), c(52.5, 67.5), lty = 2)
lines(c(-8.75, -1.25), c(72.5, 87.5), lty = 2)
lines(c(18.75, 11.75), c(52.5, 67.5), lty = 2)
lines(c(8.75, 1.25), c(72.5, 87.5), lty = 2)
@
\caption{Residual-based shadings in HSV (upper) and HCL space (lower).}
\label{fig:shadingHSVHCL}
\end{center}
\end{figure}

@@

The shading scheme of \cite{vcd:Friendly:1994} was suggested to visualize
the pattern of dependence in contingency tables, as discussed above, but
the presence (or absence) of colors in a plot also always conveys an
impression of interestingness (or un-interstingness, respectively). That is,
viewers might be tempted to interpret the absence of color in a plot
as a clue that there is no significant departure from independence. Or vice
versa, colored cells would convey the impression that there is significant
dependence. Currently, both is not true as can be seen in the left panel
of Figure~\ref{fig:shadings} which shows the mosaic display for the arthritis
data with Friendly shading. Although there is significant dependence, no 
residual exceeds an absolute value of $2$ and hence no cell is colored.
Of course, it can be argued that the shading was not designed for this purpose
and that different cut offs than $2$ and $4$ should be used here. However,
in this situation, it would be nice if such cut offs could be chosen automatically in a
data-driven way. Strategies for this are derived in the following.

The Friendly shading can be interpreted to be a visualization of the maximum
statistic $M$ from Equation~(\ref{eq:MaxAbs}) which always employs the critical values $c_\alpha$
$2$ and $4$. The problem is that it is not clear to which
significance levels $\alpha$ these critical values correspond because the distribution
of $M$ depends on the underlying contingency table. The natural solution to this
problem is to compute the critical values from the distribution of $M$ in a data-driven
way (i.e., for the table visualized) and use these instead of the hard-coded values
$2$ and $4$. In the right panel of Figure~\ref{fig:shadings} this is done for
the arthritis data by employing the critical values
$\Sexpr{round(as.vector(art_max$qdist(0.9)), digits = 2)}$ at level $\alpha = 0.1$ and
$\Sexpr{round(as.vector(art_max$qdist(0.99)), digits = 2)}$ at level $\alpha = 0.01$
(and using the diverging HCL palette) derived from the permutation distribution of $M$ for 
the arthritis data as described in Section~\ref{sec:tests}. By using these cut offs, the presence of color
in the plot is equivalent to significance (of the maximum statistic $M$) at 
level $\alpha = 0.1$ and $\alpha = 0.01$, respectively, and exactly the cells which
violate the independence hypothesis are highlighted. For the arthritis data, these are in particular the 
cells in the last column that signal that there are significantly more
treated patients and fewer placebo patients with marked improvement than would be expected
under independence.

@@

This maximum shading is already very flexible and combines visualization and inference.
However, it can only be applied when employing the maximum statistic because it is the only
aggregation functional $\lambda(\cdot)$ where a single large residual $|r_{ij}|$ exceeding
its critical value is equivalent
to a significant value of the whole test statistic $\lambda([r_{ij}])$.\footnote{This
can easily be seen from the following consideration: To obtain a level $\alpha$ test, where 
any residual $|r_{ij}|$ exceeding the critical value $c_\alpha$ gives a significant test
result, the equation $\alpha = P(\mbox{any } |r_{ij}| > c_\alpha)$ must hold. This is equivalent
to $1 - \alpha = P(\mbox{all } |r_{ij}| \le c_\alpha) = P(\max_{i,j} |r_{ij}| \le c_\alpha)$.}
Typically, applying the maximum statistic is feasible and also appropriate for exploratory 
analysis, but it would be desirable to also have a residual-based shading that can incorporate
visualization of significance when the sum-of-squares statistic $X^2$ (or any other functional
$\lambda(\cdot)$) is used. For the reasons discussed above, it is not possible to achieve this by shading
individual cells differently but can only be realized by using different colors for the
whole table. As outlined before, colorfulness is intuitively matched with interestingness,
therefore a rather natural idea is to use the fully colored palette only when the corresponding
test is significant and to use a less colorful palette if not. For the HCL scheme, the
amount of color can conveniently be controlled by varying the maximum chroma value used. For the full
colors, the maximum chroma was set to $100$ as shown in Figure~\ref{fig:shadingHSVHCL} and 
is reduced to $20$ for the non-significant palette. This palette still codes the absolute
size of the residuals by luminance (i.e., the amount of grey), uses the same neutral grey
for small `un-interesting' residuals, codes positive and negative residuals by different hues,
but gives less emphasis to the pattern by making the plot less colorful. A similar effect
can be obtained in the HSV space if the value is reduced from $1$ to $0.5$ for non-significant 
palettes (see Figure~\ref{fig:shadingHSVHCL}). As for the significant colors,
the dimensions used for creating this palette are confounded and hence the HCL scheme is clearly
preferable.

@@

To introduce the new residual-based shadings without too much overhead in
Section~\ref{sec:shadings}, we have only considered the independence problem
in 2-way tables. In this section, generalizations of these ideas to independence
problems for multi-way tables are outlined.

Mosaic displays have been emphasized in the literature to be an excellent means
of visualization for log-linear models \citep{vcd:Friendly:1999,vcd:Theus+Lauer:1999}.
Problems of interest include complete, joint or conditional independence for all of 
which the tables of expected values and residuals (again Pearson or deviance) can
be computed. \cite{vcd:Friendly:1994,vcd:Friendly:1999} shows that his residual-based 
shading scheme can directly be applied to these more complex independence models.
For inference, the most commonly used aggregation functional for the residuals
is again the sum of squares yielding the associated Pearson or likelihood
ratio statistic, respectively \citep{vcd:Agresti:2002}.

As these independence models for multi-way tables also provide the structure
required for the residual-based HCL shadings derived in Section~\ref{sec:shadings},
both the maximum shading and the sum-of-squares shading can straightforwardly be applied.
However, such independence models often additionally provide further structure that allows
decomposition of the overall model into smaller independence problems which can be exploited
both for choosing appropriate data-driven cut-offs in the shading and for selecting 
a suitable layout of mosaic or association plots. Specific strategies for the conditional
independence problem are derived in the following.

@@

How the structure of the independence problem can be exploited for 
selecting a suitable layout and shading for mosaic and association plots,
is exemplified with the conditional independence problem in 3-way tables.
For a table $[n_{ijk}]$ with underlying
theoretical probability distribution $[p_{ijk}]$, this can be formulated as
\begin{equation} \label{eq:H0ci}
H_0: \; \pi_{ij|k} \quad = \quad \pi_{i+|k} \pi_{+j|k}.
\end{equation}
where $\pi_{ij|k}$ are the conditional probabilities given the stratum $k$
with $k = 1, \dots, K$. Under the assumption of conditional independence,
we can again compute expected frequencies $[\hat n_{ijk}]$ and the corresponding 
residuals $[r_{ijk}]$. To test the conditional independence hypothesis, usually
the sum-of-squares statistic is used
\begin{equation} \label{eq:sumChisq}
\sum_{i, j, k} r_{ijk}^2 \quad = \quad \sum_k X_k^2,
\end{equation}
which is simply the sum of the individual sum-of-squares statistics $X_k^2$ in each stratum $k$.

@@

All these statistics are of type
\begin{equation} \label{eq:coindep_test}
\lambda_\mathrm{agg} (\lambda_\mathrm{indep}([r_{ijk}])),
\end{equation}
where $\lambda_\mathrm{indep}$ is a functional for assessing independence in stratum $k$
and $\lambda_\mathrm{agg}$ is a functional for aggregating over the $k = 1, \dots, K$ strata.
If the maximum is used for the latter, then identification of the strata responsible for the
dependence is possible. If additionally $\lambda_\mathrm{indep}$ is the maximum,
the corresponding cells can also be identified. Hence, the double maximum
statistic from Equation~(\ref{eq:doublemax}) is the only functional allowing for detection of
both the strata and the cells violating the conditional independence hypothesis.

However, the main purpose of the formulation of the different test statistics is not so much 
inference but their applicability to diagnostic plots via residual-based shadings. As already
discussed, it is possible for all aggregation functionals to simply use either
the significant or the non-significant shading for all cells in the contingency table---this
strategy would have to be used for the sum-of-squares statistic from Equation~(\ref{eq:sumChisq}). If
$\lambda_\mathrm{agg}$ is the maximum as in Equation~(\ref{eq:maxChisq}),
then the significant palette would only be used
in those strata in conflict with the hypothesis of conditional independence whereas the
non-significant palette would be used for the remaining strata. Finally, if both $\lambda_\mathrm{agg}$
and $\lambda_\mathrm{indep}$ are the maximum, then the same strategy as in Section~\ref{sec:shadings}
can be pursued, i.e., only the significant palette is used but with data-driven cut offs
derived from the distribution of the double maximum statistic from Equation~(\ref{eq:doublemax}).

@@

Various strategies for constructing residual-based shadings for visualizing
(conditional) independence in contingency tables via mosaic and association plots
are discussed. The shading
of \cite{vcd:Friendly:1994} is extended in two directions: the use of perceptually
uniform HCL colors and the combination of visualization and significance testing.
To achieve the former, a general guideline for constructing diverging palettes
in HCL space is introduced. The advantages of using this HCL shading scheme
instead of an HSV scheme are that the colors from this perceptually-based color space
provide uniform saturations over different hues and that the colorfulness in this
shading can be controlled independently from the other two dimensions.

To combine visualization and significance testing, two approaches are presented:
The first approach always uses a fully colored palette but relies on
data-driven cut offs such that the presence of color is equivalent to significance of the
associated maximum test. The second approach uses pre-defined cut offs (such as $2$ and $4$) but
codes the result of the significance test by using full colors only for significant results and
the same type of palette with a reduced amount of color for non-significant results.
